---
title: "Combined QUS + Carlson LQMM Analysis with all post-partum data"
output: pdf_document
---

**Comment (dgs): I added a QUS data filter for "Outcome==0", we only use full term data for the combined analysis.**

**Comment 2 (dgs): In this version I extended figures out to 100 weeks to include all the post-partum, full-term data**

```{r setup, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(splines)
library(lqmm)
library(quantreg)   # used only for the rq fallback at the bottom
```

```{r config}
# ---------------------------------------------------------------------------
# Paths -- adjust to wherever the two files live on your machine.
# ---------------------------------------------------------------------------
qus_path     <- "../../../data/V1_V7_longformat_k1.csv"
carlson_path <- "carlson_et_al_data.txt"

# ---------------------------------------------------------------------------
# Carlson has ~91k measurement-level rows; QUS has ~1k. Set this TRUE to
# collapse Carlson to one Avg_SWS per exam (ID x GA x Exam) so the two
# datasets contribute on a comparable scale to the pooled LQMM. Set FALSE to
# keep every measurement row (your original (P+MP1+MP2+M)/4 per-row approach).
# ---------------------------------------------------------------------------
collapse_carlson_per_exam <- TRUE

taus <- c(0.2, 0.5, 0.8)
```

# 1. QUS data (full data -- no PTB / FTB split)

```{r qus-prep}
qus_raw <- read_csv(qus_path, show_col_types = FALSE)

qus <- qus_raw %>%
  filter(mPTB != "1") %>%                       # same clinical filter as before
#  filter(!Visit_ID %in% c(5, 6, 7)) %>%         # drop late visits as before
  filter(Outcome == "0") %>%              ##### full term only (-dgs)
  mutate(GA = GA * 39) %>%                       # convert GA to weeks
  transmute(
    Participant_ID = paste0("QUS_", Participant_ID),  # prefix to avoid ID collision
    GA             = GA,
    Avg_SWS        = SWS,                              # rename SWS -> Avg_SWS
    source         = "QUS"
  ) %>%
  filter(is.finite(GA), is.finite(Avg_SWS), !is.na(Participant_ID))

nrow(qus)
range(qus$GA)
```

# 2. Carlson data (average across P, MP1, MP2, M)

```{r carlson-prep}
carlson_raw <- read.table(carlson_path, header = TRUE, sep = "\t")

carlson <- carlson_raw %>%
  mutate(Avg_SWS = (P + MP1 + MP2 + M) / 4)     # average the 4 positions per row

# Optionally collapse to one value per exam so Carlson doesn't dominate the pool
if (collapse_carlson_per_exam) {
  carlson <- carlson %>%
    group_by(ID, Exam, GA) %>%
    summarise(Avg_SWS = mean(Avg_SWS, na.rm = TRUE), .groups = "drop")
}

carlson <- carlson %>%
  transmute(
    Participant_ID = paste0("CARL_", ID),         # prefix to avoid ID collision
    GA             = GA,                           # Carlson GA already in weeks
    Avg_SWS        = Avg_SWS,
    source         = "Carlson"
  ) %>%
  filter(is.finite(GA), is.finite(Avg_SWS), !is.na(Participant_ID))

nrow(carlson)
range(carlson$GA)
```

# 3. Combine on GA and Avg_SWS

```{r combine}
combined <- bind_rows(qus, carlson) %>%
  mutate(Participant_ID = as.factor(Participant_ID)) %>%
  arrange(Participant_ID, GA)

table(combined$source)
range(combined$GA)
range(combined$Avg_SWS)
```

# 4. Spline basis, 4-week bins, and prediction grid

```{r basis-bins-grid}
# ---- spline basis on the pooled GA -- KEEP it so prediction uses identical knots
ga_ns    <- ns(combined$GA, knots = c(16, 20, 24, 31, 36, 40, 44, 48, 64))
basis_df <- as.data.frame(ga_ns)
names(basis_df) <- paste0("ns", seq_len(ncol(basis_df)))
combined <- bind_cols(combined, basis_df)

# ---- 4-week bins FIRST, so GA_mid exists before we build the grid ----
brks     <- seq(6, 96, by = 5)
mids     <- head(brks, -1) + 2
bin_labs <- paste0("[", head(brks, -1), ",", tail(brks, -1), ")")
bin_labs[length(bin_labs)] <- sub(")$", "]", bin_labs[length(bin_labs)])

combined <- combined %>%
  mutate(
    GA_bin = cut(GA, breaks = brks, right = FALSE, include.lowest = TRUE),
    GA_mid = brks[as.integer(GA_bin)] + 2
  ) %>%
  filter(!is.na(GA_mid))

# ---- grid extended to the last occupied box midpoint ----
grid_ga <- seq(min(combined$GA), max(combined$GA_mid), length.out = 300)
grid_X  <- cbind(Intercept = 1, predict(ga_ns, grid_ga))

table(combined$GA_bin)
```

# 5. Combined LQMM -- Avg_SWS (linear scale)

```{r lqmm-linear}
fixed_form <- as.formula(paste("Avg_SWS ~", paste(names(basis_df), collapse = " + ")))

preds <- bind_rows(lapply(taus, function(t) {
  fit  <- lqmm(
    fixed   = fixed_form,
    random  = ~ 1,
    group   = Participant_ID,
    tau     = t,
    data    = combined,
    control = lqmmControl(LP_max_iter = 1000)
  )
  beta <- as.numeric(coef(fit))
  data.frame(GA = grid_ga, fit = as.numeric(grid_X %*% beta),
             tau = factor(paste0("tau= ", t)))
}))
```

```{r plot-linear, fig.width = 9, fig.height = 6}
y_top <- quantile(combined$Avg_SWS, 0.995, na.rm = TRUE) * 1.05   # robust upper limit

combined_plot <- ggplot(combined, aes(x = GA_mid, y = Avg_SWS, group = GA_mid)) +
  geom_boxplot(outlier.alpha = 0.1, fill = "grey85", width = 3) +
  geom_line(
    data = preds,
    aes(x = GA, y = fit, color = tau, group = tau),
    inherit.aes = FALSE,
    linewidth = 1.3
  ) +
  scale_color_manual(values = c(
    "tau= 0.2" = "#F8766D",
    "tau= 0.5" = "#00BA38",
    "tau= 0.8" = "#619CFF"
  )) +
  scale_x_continuous(breaks = mids, labels = bin_labs) +
  labs(
    title    = "Avg_SWS vs GA (Combined QUS + Carlson)",
    subtitle = "Grey: 4-week Boxplots | Colored Lines: LQMM Population Quantiles",
    x        = "Gestational Age (weeks)",
    y        = "Avg_SWS",
    color    = "Population Quantiles"
  ) +
  coord_cartesian(xlim = c(5,100), ylim = c(0, y_top)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

combined_plot
```

# 6. Combined LQMM -- log(Avg_SWS)

```{r lqmm-log}
combined_log <- combined %>%
  filter(Avg_SWS > 0) %>%
  mutate(log_SWS = log(Avg_SWS))

# grid can change slightly if the Avg_SWS > 0 filter dropped the last bin
grid_ga_log <- seq(min(combined_log$GA), max(combined_log$GA_mid), length.out = 300)
grid_X_log  <- cbind(Intercept = 1, predict(ga_ns, grid_ga_log))

log_form <- as.formula(paste("log_SWS ~", paste(names(basis_df), collapse = " + ")))

preds_log <- bind_rows(lapply(taus, function(t) {
  fit  <- lqmm(
    fixed   = log_form,
    random  = ~ 1,
    group   = Participant_ID,
    tau     = t,
    data    = combined_log,
    control = lqmmControl(LP_max_iter = 1000)
  )
  beta <- as.numeric(coef(fit))
  data.frame(GA = grid_ga_log, fit = as.numeric(grid_X_log %*% beta),
             tau = factor(paste0("tau= ", t)))
}))
```

```{r plot-log, fig.width = 9, fig.height = 6}
y_lo <- min(combined_log$log_SWS, na.rm = TRUE)
y_hi <- quantile(combined_log$log_SWS, 0.995, na.rm = TRUE)

combined_plot_log <- ggplot(combined_log, aes(x = GA_mid, y = log_SWS, group = GA_mid)) +
  geom_boxplot(outlier.alpha = 0.1, fill = "grey85", width = 3) +
  geom_line(
    data = preds_log,
    aes(x = GA, y = fit, color = tau, group = tau),
    inherit.aes = FALSE,
    linewidth = 1.3
  ) +
  scale_color_manual(values = c(
    "tau= 0.2" = "#F8766D",
    "tau= 0.5" = "#00BA38",
    "tau= 0.8" = "#619CFF"
  )) +
  scale_x_continuous(breaks = mids, labels = bin_labs) +
  labs(
    title    = "log(Avg_SWS) vs GA (Combined QUS + Carlson)",
    subtitle = "Grey: 4-week Boxplots | Colored Lines: LQMM Population Quantiles",
    x        = "Gestational Age (weeks)",
    y        = "log(Avg_SWS)",
    color    = "Population Quantiles"
  ) +
  coord_cartesian(xlim = c(5, 100), ylim = c(y_lo, y_hi)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

combined_plot_log
```

# 7. Optional: see how each dataset contributes

```{r}
# Same combined LQMM curves, but boxplots split by source so you can see which
# bins are QUS, which are Carlson, and where they overlap.
ggplot(combined, aes(x = GA_mid, y = Avg_SWS, group = interaction(GA_mid, source),
                     fill = source)) +
  geom_boxplot(outlier.alpha = 0.1, width = 3, position = position_dodge(width = 2)) +
  geom_line(
    data = preds,
    aes(x = GA, y = fit, color = tau, group = tau),
    inherit.aes = FALSE,
    linewidth = 1.3
  ) +
  scale_fill_manual(values = c("QUS" = "lightblue", "Carlson" = "orange")) +
  scale_color_manual(values = c(
    "tau= 0.2" = "#F8766D",
    "tau= 0.5" = "#00BA38",
    "tau= 0.8" = "#619CFF"
  )) +
  scale_x_continuous(breaks = mids, labels = bin_labs) +
  labs(
    title    = "Avg_SWS vs GA by source (Combined LQMM curves overlaid)",
    x        = "Gestational Age (weeks)",
    y        = "Avg_SWS",
    fill     = "Source",
    color    = "Population Quantiles"
  ) +
  coord_cartesian(xlim = c(6, 96), ylim = c(0, y_top)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Same combined log-scale LQMM curves, but boxplots split by source so you can
# see which bins are QUS, which are Carlson, and where they overlap.
y_lo <- min(combined_log$log_SWS, na.rm = TRUE)
y_hi <- quantile(combined_log$log_SWS, 0.995, na.rm = TRUE)

ggplot(combined_log, aes(x = GA_mid, y = log_SWS,
                         group = interaction(GA_mid, source),
                         fill = source)) +
  geom_boxplot(outlier.alpha = 0.1, width = 3, position = position_dodge(width = 2)) +
  geom_line(
    data = preds_log,
    aes(x = GA, y = fit, color = tau, group = tau),
    inherit.aes = FALSE,
    linewidth = 1.3
  ) +
  scale_fill_manual(values = c("QUS" = "lightblue", "Carlson" = "orange")) +
  scale_color_manual(values = c(
    "tau= 0.2" = "#F8766D",
    "tau= 0.5" = "#00BA38",
    "tau= 0.8" = "#619CFF"
  )) +
  scale_x_continuous(breaks = mids, labels = bin_labs) +
  labs(
    title    = "log(Avg_SWS) vs GA by source (Combined LQMM curves overlaid)",
    x        = "Gestational Age (weeks)",
    y        = "log(Avg_SWS)",
    fill     = "Source",
    color    = "Population Quantiles"
  ) +
  coord_cartesian(xlim = c(8, 100), ylim = c(y_lo, y_hi)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
